Gaze-dependent evidence accumulation predicts multi-alternative risky choice behaviour

### 2021-02-05 manuscript/main.tex
renamed figures, separate reference list for supplement, added draft for contributions and acknowledgments

% Color links
\usepackage{hyperref}
\hypersetup{
    colorlinks,
    linkcolor={red!50!black},                           % Color of within-text links (e.g., to a Figure)
    citecolor={blue!50!black},                          % Color of citation links (within text, to Bibliography)
    urlcolor={blue!50!black}                            % Color of URLs
}

@@

Crucially, choices in these multi-alternative, multi-attribute settings pose a challenge to many traditional models of risky choice, including Expected Utility Theory \parencite[EU][]{vonneumann1947TheoryGamesEconomic} and Prospect Theory \parencite{kahneman1979ProspectTheoryAnalysis,tversky1992AdvancesProspectTheory}, which obey basic axioms of rational choice like independence of irrelevant alternatives \parencite[IIA;][]{luce1959IndividualChoiceBehavior}. Briefly, IIA asserts that the preference between two alternatives should not depend on other alternatives. At least three context effects – the attraction \parencite{huber1982AddingAsymmetricallyDominated}, compromise \parencite{simonson1989ChoiceBasedReasons}, and similarity \parencite{tversky1972EliminationAspectsTheory} effects – show that IIA is frequently violated: The attraction effect describes an increase in preference for an alternative following the addition of a similar, but slightly inferior alternative. The compromise effect describes an increase in preference for an alternative after a third alternative was added that makes it appear as intermediate. The similarity effect predicts that adding a third option that is similar to one of the original options, and equally appealing, will increase relative preference for the other, dissimilar alternative. While these effects are predominantly investigated using consumer goods, some studies also found them to affect choices between risky gambles \parencite{huber1982AddingAsymmetricallyDominated,mohr2017AttractionEffectRisky,soltani2012RangeNormalizationModelContextDependent,tversky1972EliminationAspectsTheory,wedell1991DistinguishingModelsContextually}.

@@

% Figure 1: Choice task and attribute space
\begin{SCfigure}[0.8]
\caption{\textbf{Choice task and attribute space of stimuli.} \textbf{(a)} Choice task. Trials started with a central fixation cross for 1.5 s. Next, three gambles were presented. Participants made a choice by button press, without time limit. Eye movements were recorded during the choice phase. Finally, brief feedback on the choice (but not on the gamble outcome) was displayed. \textbf{(b)} Attribute space. Each gamble is described by two attributes: Probability $p$ and outcome $m$. The core options $A$ and $B$ were presented in every trial, along with one of four decoy alternatives: Compromise decoys ($C_A$ or $C_B$) and asymmetrically dominated decoys ($D_A$ or $D_B$) are expected to elicit compromise and attraction effects, respectively. Dashed lines indicate possible stimulus placements. \textbf{(c)} Exact positions for $C_B$, $A$ and $C_A$ were calibrated for pairwise indifference in separate binary choice estimation blocks for each individual. Differences in outcome between neighbouring alternatives was not less than 2 EUR. Dominated decoys $D_A$ and $D_B$ were always 2\% and 1 EUR worse than $A$ and $B$ respectively.}
\includegraphics[scale=1]{../figures/1-task_stimuli.pdf}
\hspace{0.5cm}
\label{fig:task}
\end{SCfigure}

In our experiment, 40 participants made repeated decisions between three all-or-nothing gambles, each described by a probability $p$ to win an outcome $m$ and nothing otherwise (Figure \ref{fig:task}a). We recorded participants' eye movements during the task with an eye tracker (see \nameref{sec:methods} for details). Participants were instructed that after completing the task, their chosen gamble from one randomly determined trial would be played out for an additional bonus payment. In contrast to hypothetical choices between consumer goods described on attribute dimensions like quality and price, risky gamble stimuli offer a high amount of control over their attributes and a straightforward way to incentivise choices. The gamble stimuli were designed to elicit attraction and compromise effects and were individually tailored to account for each participant's risk preferences (Figure \ref{fig:task}b). Participants performed three pairs of indifference-estimation and experimental blocks, for a total of 225 experimental trials (see \nameref{sec:methods} for additional details on the experimental procedure and gamble stimuli). In indifference-estimation blocks, participants made repeated choices between pairs of gambles with different probabilities (Figure \ref{fig:task}c). Gamble outcomes $m$ were adjusted according to participants' choices so that four approximately equally preferred gambles $C_B$, $B$, $A$, and $C_A$ were constructed with winning probabilities $p$ = 55\%, 65\%, 75\%, and 85\%. Additionally, two asymmetrically dominated decoys $D_A$ and $D_B$ were defined to be 2\% and 1 EUR worse than gambles $A$ and $B$, respectively. Following this indifference estimation, participants performed an experimental block of 75 ternary choice trials, 32 of which were compromise trials balanced with respect to the target option (i.e., 16 choice sets \{$A$, $B$, $C_A$\}, 16 choice sets \{$A$, $B$, $C_B$\}), 32 attraction trials (16 choice sets \{$A$, $B$ ,$D_A$\}, 16 choice sets \{$A$, $B$, $D_B$\}) and 11 distractor trials showing randomly created options with expected value of 10 EUR and low (5–33\%), medium (34–64\%) and high (65–95\%) probability $p$. Uniform noise was added to all outcomes $m$ ($\pm$0 EUR, +1 EUR) and probabilities $p$ (–3\%, $\pm$0\%, +3\%) in each trial. Asymmetrically dominated decoys $D_A$ and $D_B$ received the same noise as the target option to preserve the dominance relation.

@@

We first analysed participants' choice behaviour to test whether their choices could be influenced by the set of available alternatives (see Table \ref{tab:choicefreqs} and Figure \ref{fig:cfx}). Participants rarely chose the dominated decoys in attraction trials (mean $\pm$ s.d. = 0.02 $\pm$ 0.02 of attraction trials), indicating that participants understood the dominance relationships among the stimuli. In compromise trials, participants chose (non-dominated) decoy alternatives more frequently (mean $\pm$ s.d. = 0.27 $\pm$ 0.16). In particular, the decoy with the highest possible outcome $C_B$ ($p \approx$ 55\%, $m \approx$ 18 EUR) was chosen most frequently when it was available. Note that decoy choices in compromise trials are expected, as extreme decoys were specifically calibrated to be approximately equally preferred to the core options.

@@

We find evidence for the attraction effect: The mean RST in attraction trials differed meaningfully from 0.5 (mean RST = 0.56, HDI$_{95}$ = [0.51, 0.60], mean $d$ = 0.49, HDI$_{95}$ = [0.10, 0.80]). 25 of 40 (62.5\%) participants had an RST above 0.5 in attraction trials. Notably, similar to previous work \parencite[e.g.,][]{trueblood2012MultialternativeContextEffects} a subgroup of participants (9 of 40, 22.5\%) showed particularly strong attraction effects with individual RSTs above 0.7. We could not, however, find evidence that these individuals used the dominance relationship as a simplifying choice rule (see \nameref{sec:supplement}). To allow comparisons with other studies that quantified context effects as the difference between choice shares between targets and competitors, we also report these differences: In attraction trials, the average difference was 0.12 (HDI$_{95}$ = [0.01, 0.20], mean $d$ = 0.48, HDI$_{95}$ = [0.10, 0.85], Figure \ref{fig:cfx}b).

We only found weak evidence for the compromise effect using the gamble stimuli: The mean RST in compromise trials was 0.53, but its estimated HDI$_{95}$ did not exclude 0.5 (HDI$_{95}$ = [0.49, 0.57], 91.1\% of posterior mass above 0.5, mean $d$ = 0.23, HDI$_{95}$ = [-0.11, 0.59]). 26 of 40 (65\%) participants showed an RST above 0.5 in compromise trials. The mean difference between choice shares for targets and competitors was 0.05 (HDI$_{95}$ = [-0.01, 0.11], mean $d$ = 0.29, HDI$_{95}$ = [-0.04, 0.62], 95.8 \% of posterior mass above 0; Figure \ref{fig:cfx}e). These results are similar to the marginal effects obtained using perceptual stimuli in previous work \parencite{trueblood2015FragileNatureContextual,trueblood2013NotJustConsumers}.

@@

% Figure 2: Context effects in choices and relative dwell times
\begin{figure}[bt!]
\begin{centering}
\includegraphics[scale=1]{../figures/2-context-effects.pdf}
\caption{\textbf{Context effects are present in choices and relative dwell time.} Participants choices were influenced by asymmetrically dominated decoys and, to a lesser extent, by extreme compromise-making decoys. \textbf{(a, d)} Ternary plots of individual relative choice frequencies for target (lower left, teal), competitor (top, pink), and decoy (lower right, yellow) alternatives in attraction \textbf{(a)} and compromise \textbf{(d)} trials. Each dot represents one participant. The position on the simplex indicates relative choice frequencies for alternatives. Straight lines from the centre indicate equal frequencies for two alternatives. The red "x" indicates the group average. \textbf{(b, e)} Relative choice frequencies in attraction \textbf{(b)} and compromise \textbf{(e)} trials. In attraction trials, some participants strongly favoured the target alternative and almost no decoy choices were made. While target alternatives are still chosen more frequently than competitors in compromise trials, the effect is less pronounced, and extreme decoys are still chosen frequently. \textbf{(c, f)} Relative dwell time towards alternatives. In both, attraction \textbf{(c)} and compromise \textbf{(f)} trials, target alternatives received greater relative dwell times than competitors. $d$ denotes Cohen's $d$ from paired BEST analysis with HDI$_{95}$ given in brackets. Violin plots show kernel density estimates of distributions of individual values. Box plots mark lower and upper quartiles and median. Whiskers extend from first and last datum within 1.5 times the interquartile range from lower and upper quartiles, respectively. Values outside this range are indicated by open circles.}
\label{fig:cfx}
\end{centering}
\end{figure}

@@

Next, we tested whether participants' eye movements were affected by the set of available alternatives, similar to their choices. We therefore computed each alternative's relative dwell time in a trial by summing the duration of all fixations durations towards it, and normalizing to the sum of all fixations in the trial. Both patterns of context effects of choice behaviour were present in the relative dwell data: Target options received greater relative dwell time than competitors in attraction (mean difference = 0.014, HDI$_{95}$ = [0.004, 0.024], mean $d$ = 0.53, HDI$_{95}$  = [0.15, 0.93], Figure \ref{fig:cfx}c) and compromise trials (mean difference = 0.022, HDI$_{95}$  = [0.014, 0.030], mean $d$ = 0.90, HDI$_{95}$  = [0.53, 1.30], Figure \ref{fig:cfx}f). Targets and competitors also received greater relative dwell time than decoys in both trial types.

To better understand participants' eye movements over the course of the trial, we performed additional exploratory analyses of gaze location and transitions. We found an increasing association between gaze and choice over the trial, and longer gaze towards targets, even accounting for choice. Information search occurred both within and between alternatives, with slightly more transitions within alternatives. We refer to the \nameref{sec:supplement} for additional details on these analyses.

@@

% Figure 3: Model comparison results, predicted gaze-choice association RST prediction of winning model
\begin{figure}[bt!]
\begin{centering}
\includegraphics[scale=1]{../figures/3-model-comparison.pdf}
\caption{\textbf{Model comparison and predictions.} \textbf{(a)} The gaze-dependent leaky accumulator (GLA) provided the best fit (lowest mean BIC) across participants, followed by the dynamic gaze baseline model, MDFT, EU, and the static gaze baseline model. The dashed line indicates the BIC of the random choice baseline model. Violin plots show kernel density estimates of distributions of individual values. Box plots mark lower and upper quartiles and median. Whiskers extend from first and last datum within 1.5 times the interquartile range from lower and upper quartiles, respectively. Values outside this range are indicated by open circles. \textbf{(b)} The GLA fitted most (36 of 40, 90\%) participants best, with a protected exceedance probability of 1 (inset). \textbf{(c-h)} Observed and model-predicted probability of choosing the target \textbf{(c, d)}, competitor \textbf{(e, f)}, or decoy \textbf{(g, h)} alternatives in attraction \textbf{(c, e, g)} and compromise trials \textbf{(d, f, h)} as a function of relative dwell time advantage. Relative dwell time advantage was computed as relative dwell time towards an alternative minus the mean relative dwell time to all other alternatives. White bars and error bars show mean $\pm$ s.e. observed data from even-numbered trials. Model predictions (coloured lines) are based on 50 simulations of each odd-numbered trial. \textbf{(i, j)} Observed and predicted RST of the best-fitting GLA for attraction \textbf{(i)} and compromise \textbf{(j)} trials. Each circle represents one participant. The winning model’s predicted context effect sizes correlated significantly with the observed ones. Strong context effects, however, were underestimated, as indicated by the reduced slopes.}
\label{fig:modelcomparison}
\end{centering}
\end{figure}

@@

where all $X_i(0)=0$. Similar to the aDDM, the $\theta$ parameter ($0 \le \theta \le 1$) controls discounting of unattended alternative utilities. The $\lambda$ parameter ($0 \le \lambda \le 1$) controls the strength of the accumulation leak. We computed choice probabilities from this model by applying the soft-max choice rule (Equation \ref{eq:softmax}) over the final accumulator values $X_i^{GLA}$.

@@

Across the group, all models performed better than the random baseline model (Figure \ref{fig:modelcomparison}a). The GLA had the lowest BIC (mean $\pm$ s.d. = 230.63 $\pm$ 66.80), followed by the dynamic gaze baseline model GB$_{\text{dyn}}$ (mean $\pm$ s.d. = 302.46 $\pm$ 86.15), MDFT (mean $\pm$ s.d. = 345.77 $\pm$ 82.44), EU (mean $\pm$ s.d. = 360.57 $\pm$ 72.56) and the static gaze baseline GB$_{\text{stat}}$ (mean $\pm$ s.d. = 414.81 $\pm$ 36.81). Simulating choices from the models using maximum-likelihood estimates, the proportion of correctly predicted choices was 74.2\% for GLA, 62.8\% for GB$_{\text{dyn}}$, 57.9\% for MDFT, 53.8\% for EU and 45.7\% for GB$_{\text{stat}}$ (see Supplementary Figure \ref{fig:model-pred-probs} for distributions of model-predicted choice probabilities). Note, that target and competitor options (and decoys in compromise trials) were designed to be closely matched in value, resulting in trials with high choice difficulty, limiting overall model performance. On the individual level, based on lowest BIC scores, the majority (36 of 40, 90\%) of participants were best described by GLA. Three participants (7.5\%) were best described by MDFT, and one by the dynamic gaze-baseline model (Figure \ref{fig:modelcomparison}b). Protected exceedance probabilities \parencite{rigoux2014BayesianModelSelection}, which measure the likelihood that model is more frequent than all others, unambiguously identified GLA as the most likely model (pXP$_{\text{GLA}}$ = 1, Figure \ref{fig:modelcomparison}b inset).

The estimates for GLA’s gaze-discount parameter $\theta$, which describes the degree to which alternative’s values are attenuated while not fixated, indicate that participants exhibited a moderate gaze-discount on average: $\theta$ estimates ranged from 0.13 (strong gaze-discount) to 0.95 (almost no gaze-discount), with mean $\pm$ s.d. = 0.69 $\pm$ 0.18. Estimates for the leak parameter ranged from 0.08 (almost no leak) to 0.65 (moderately strong leak), with mean $\pm$ s.d. = 0.29 $\pm$ 0.20. Individual parameter estimates of all GLA parameters are plotted in Supplemental Figure \ref{fig:gla-estimates} and summarised in Supplemental Table \ref{tab:gla-estimates}.

Given that MDFT outperforms utility-based models when choices are influenced by context \parencite{berkowitsch2014RigorouslyTestingMultialternative}, we tested whether model fit of MDFT (relative to GLA) was higher for participants with higher RST. In other words: Does MDFT perform better for stronger context effects? Overall, 7 out of 9 participants with an RST above 0.7 in attraction trials were clearly best described by GLA. While we found that the relative superiority of GLA over MDFT decreased with strong attraction effects, this was not the case for compromise trials (Supplemental Figure \ref{fig:dbic-rst}), suggesting that some features of MDFT might help capture strong attraction effects.

As an additional indicator of model fit, we tested whether the models could quantitatively reproduce the observed positive association of gaze and choice (Supplemental Figures \ref{fig:dwell-data-timebinned}d, j, \ref{fig:dwell-regression-weights}). Specifically, following previous work \parencite{krajbich2010VisualFixationsComputation,krajbich2011MultialternativeDriftdiffusionModel}, we inspected the model-predicted probability of choosing an alternative as a function of its relative dwell time advantage: the difference in relative dwell time towards an alternative minus the mean relative dwell time to other alternatives. The probability of choosing an alternative generally increased with its relative dwell time advantage (Figure \ref{fig:modelcomparison}c-h), except for dominated decoys in attraction trials, which were not chosen even if looked at longer than other alternatives (Figure \ref{fig:modelcomparison}g).

All gaze-dependent models were able to capture this positive association. Note, however, that they also predicted choices of dominated decoys in attraction trials, if decoys had a large dwell time advantage (Figure 3g). In this case, GLA performed better than GB$_{\text{dyn}}$ and GB$_{\text{stat}}$, as it predicted fewer decoy choices. However, MDFT and EU generally could not capture the empirical association of gaze and choice; they predicted too many choices of alternatives with negative dwell time advantage, and too few choices of alternatives with positive dwell time advantage. 
Nor did they predict choices of dominated decoys, even if the decoy was looked at longer.

Finally, we investigated the ability of the best-performing model to predict individual differences in context effect strengths. Therefore, we predicted choices from the fitted GLA model and correlated the resulting RST with the observed data. Predicted RST correlated significantly with observed ones in attraction ($r$ = 0.76, HDI$_{95}$ = [0.62, 0.90], Figure \ref{fig:modelcomparison}i) and compromise ($r$ = 0.82, HDI$_{95}$ = [0.72, 0.92], Figure \ref{fig:modelcomparison}j) trials. However, the model underestimates large deviations from RST = 0.5, suggesting that its gaze-discount mechanism can capture the qualitative pattern of context effects but not their expression in participants with extreme RSTs.

>>>>

% Color links
\usepackage{hyperref}
\hypersetup{
    colorlinks,
    linkcolor={red!50!black},                           % Color of within-text links (e.g., to a Fig.)
    citecolor={blue!50!black},                          % Color of citation links (within text, to Bibliography)
    urlcolor={blue!50!black}                            % Color of URLs
}

@@

Crucially, choices in these multi-alternative, multi-attribute settings pose a challenge to many traditional models of risky choice, including Expected Utility Theory \parencite[EU;][]{vonneumann1947TheoryGamesEconomic} and Prospect Theory \parencite{kahneman1979ProspectTheoryAnalysis,tversky1992AdvancesProspectTheory}, which obey basic axioms of rational choice like independence of irrelevant alternatives \parencite[IIA;][]{luce1959IndividualChoiceBehavior}. Briefly, IIA asserts that the preference between two alternatives should not depend on other alternatives. At least three context effects – the attraction \parencite{huber1982AddingAsymmetricallyDominated}, compromise \parencite{simonson1989ChoiceBasedReasons}, and similarity \parencite{tversky1972EliminationAspectsTheory} effects – show that IIA is frequently violated: The attraction effect describes an increase in preference for an alternative following the addition of a similar, but slightly inferior alternative. The compromise effect describes an increase in preference for an alternative after a third alternative was added that makes it appear as intermediate. The similarity effect predicts that adding a third option that is similar to one of the original options, and equally appealing, will increase relative preference for the other, dissimilar alternative. While these effects are predominantly investigated using consumer goods, some studies also found them to affect choices between risky gambles \parencite{huber1982AddingAsymmetricallyDominated,mohr2017AttractionEffectRisky,soltani2012RangeNormalizationModelContextDependent,tversky1972EliminationAspectsTheory,wedell1991DistinguishingModelsContextually}.

@@

% Fig. 1: Choice task and attribute space
\begin{SCfigure}[0.8]
\caption{\textbf{Choice task and attribute space of stimuli.} \textbf{(a)} Choice task. Trials started with a central fixation cross for 1.5 s. Next, three gambles were presented. Participants made a choice by button press, without time limit. Eye movements were recorded during the choice phase. Finally, brief feedback on the choice (but not on the gamble outcome) was displayed. \textbf{(b)} Attribute space. Each gamble is described by two attributes: Probability $p$ and outcome $m$. The core options $A$ and $B$ were presented in every trial, along with one of four decoy alternatives: Compromise decoys ($C_A$ or $C_B$) and asymmetrically dominated decoys ($D_A$ or $D_B$) are expected to elicit compromise and attraction effects, respectively. Dashed lines indicate possible stimulus placements. \textbf{(c)} Exact positions for $C_B$, $A$ and $C_A$ were calibrated for pairwise indifference in separate binary choice estimation blocks for each individual. Differences in outcome between neighbouring alternatives was not less than 2 EUR. Dominated decoys $D_A$ and $D_B$ were always 2\% and 1 EUR worse than $A$ and $B$ respectively.}
\includegraphics[scale=1]{../figures/1-task_stimuli.pdf}
\hspace{0.5cm}
\label{fig:task}
\end{SCfigure}

In our experiment, 40 participants made repeated decisions between three all-or-nothing gambles, each described by a probability $p$ to win an outcome $m$ and nothing otherwise (Fig. \ref{fig:task}a). We recorded participants' eye movements during the task with an eye tracker (see \nameref{sec:methods} for details). Participants were instructed that after completing the task, their chosen gamble from one randomly determined trial would be played out for an additional bonus payment. In contrast to hypothetical choices between consumer goods described on attribute dimensions like quality and price, risky gamble stimuli offer a high amount of control over their attributes and a straightforward way to incentivise choices. The gamble stimuli were designed to elicit attraction and compromise effects and were individually tailored to account for each participant's risk preferences (Fig. \ref{fig:task}b). Participants performed three pairs of indifference-estimation and experimental blocks, for a total of 225 experimental trials (see \nameref{sec:methods} for additional details on the experimental procedure and gamble stimuli). In indifference-estimation blocks, participants made repeated choices between pairs of gambles with different probabilities (Fig. \ref{fig:task}c). Gamble outcomes $m$ were adjusted according to participants' choices so that four approximately equally preferred gambles $C_B$, $B$, $A$, and $C_A$ were constructed with winning probabilities $p$ = 55\%, 65\%, 75\%, and 85\%. Additionally, two asymmetrically dominated decoys $D_A$ and $D_B$ were defined to be 2\% and 1 EUR worse than gambles $A$ and $B$, respectively. Following this indifference estimation, participants performed an experimental block of 75 ternary choice trials, 32 of which were compromise trials balanced with respect to the target option (i.e., 16 choice sets \{$A$, $B$, $C_A$\}, 16 choice sets \{$A$, $B$, $C_B$\}), 32 attraction trials (16 choice sets \{$A$, $B$ ,$D_A$\}, 16 choice sets \{$A$, $B$, $D_B$\}) and 11 distractor trials showing randomly created options with expected value of 10 EUR and low (5–33\%), medium (34–64\%) and high (65–95\%) probability $p$. Uniform noise was added to all outcomes $m$ ($\pm$0 EUR, +1 EUR) and probabilities $p$ (–3\%, $\pm$0\%, +3\%) in each trial. Asymmetrically dominated decoys $D_A$ and $D_B$ received the same noise as the target option to preserve the dominance relation.

@@

We first analysed participants' choice behaviour to test whether their choices could be influenced by the set of available alternatives (see Table \ref{tab:choicefreqs} and Fig. \ref{fig:cfx}). Participants rarely chose the dominated decoys in attraction trials (mean $\pm$ s.d. = 0.02 $\pm$ 0.02 of attraction trials), indicating that participants understood the dominance relationships among the stimuli. In compromise trials, participants chose (non-dominated) decoy alternatives more frequently (mean $\pm$ s.d. = 0.27 $\pm$ 0.16). In particular, the decoy with the highest possible outcome $C_B$ ($p \approx$ 55\%, $m \approx$ 18 EUR) was chosen most frequently when it was available. Note that decoy choices in compromise trials are expected, as extreme decoys were specifically calibrated to be approximately equally preferred to the core options.

@@

We find evidence for the attraction effect: The mean RST in attraction trials differed meaningfully from 0.5 (mean RST = 0.56, HDI$_{95}$ = [0.51, 0.60], mean $d$ = 0.49, HDI$_{95}$ = [0.10, 0.80]). 25 of 40 (62.5\%) participants had an RST above 0.5 in attraction trials. Notably, similar to previous work \parencite[e.g.,][]{trueblood2012MultialternativeContextEffects} a subgroup of participants (9 of 40, 22.5\%) showed particularly strong attraction effects with individual RSTs above 0.7. We could not, however, find evidence that these individuals used the dominance relationship as a simplifying choice rule (see \nameref{sup:ade-choicerule}). To allow comparisons with other studies that quantified context effects as the difference between choice shares between targets and competitors, we also report these differences: In attraction trials, the average difference was 0.12 (HDI$_{95}$ = [0.01, 0.20], mean $d$ = 0.48, HDI$_{95}$ = [0.10, 0.85], Fig. \ref{fig:cfx}b).

We only found weak evidence for the compromise effect using the gamble stimuli: The mean RST in compromise trials was 0.53, but its estimated HDI$_{95}$ did not exclude 0.5 (HDI$_{95}$ = [0.49, 0.57], 91.1\% of posterior mass above 0.5, mean $d$ = 0.23, HDI$_{95}$ = [-0.11, 0.59]). 26 of 40 (65\%) participants showed an RST above 0.5 in compromise trials. The mean difference between choice shares for targets and competitors was 0.05 (HDI$_{95}$ = [-0.01, 0.11], mean $d$ = 0.29, HDI$_{95}$ = [-0.04, 0.62], 95.8 \% of posterior mass above 0; Fig. \ref{fig:cfx}e). These results are similar to the marginal effects obtained using perceptual stimuli in previous work \parencite{trueblood2015FragileNatureContextual,trueblood2013NotJustConsumers}.

@@

% Fig. 2: Context effects in choices and relative dwell times
\begin{figure}[bt!]
\begin{centering}
\includegraphics[scale=1]{../figures/2-context-effects.pdf}
\caption{\textbf{Context effects are present in choices and relative dwell time.} Participants choices were influenced by asymmetrically dominated decoys and, to a lesser extent, by extreme compromise-making decoys. \textbf{(a, d)} Ternary plots of individual relative choice frequencies for target (lower left, teal), competitor (top, pink), and decoy (lower right, yellow) alternatives in attraction \textbf{(a)} and compromise \textbf{(d)} trials. Each dot represents one participant. The position on the simplex indicates relative choice frequencies for alternatives. Straight lines from the centre indicate equal frequencies for two alternatives. The red "x" indicates the group average. \textbf{(b, e)} Relative choice frequencies in attraction \textbf{(b)} and compromise \textbf{(e)} trials. In attraction trials, some participants strongly favoured the target alternative and almost no decoy choices were made. While target alternatives are still chosen more frequently than competitors in compromise trials, the effect is less pronounced, and extreme decoys are still chosen frequently. \textbf{(c, f)} Relative dwell time towards alternatives. In both, attraction \textbf{(c)} and compromise \textbf{(f)} trials, target alternatives received greater relative dwell times than competitors. $d$ denotes Cohen's $d$ from paired BEST analysis with HDI$_{95}$ given in brackets. Violin plots show kernel density estimates of distributions of individual values. Box plots mark lower and upper quartiles and median. Whiskers extend from first and last datum within 1.5 times the interquartile range from lower and upper quartiles, respectively. Values outside this range are indicated by open circles.}
\label{fig:cfx}
\end{centering}
\end{figure}

@@

Next, we tested whether participants' eye movements were affected by the set of available alternatives, similar to their choices. We therefore computed each alternative's relative dwell time in a trial by summing the duration of all fixations durations towards it, and normalizing to the sum of all fixations in the trial. Both patterns of context effects of choice behaviour were present in the relative dwell data: Target options received greater relative dwell time than competitors in attraction (mean difference = 0.014, HDI$_{95}$ = [0.004, 0.024], mean $d$ = 0.53, HDI$_{95}$  = [0.15, 0.93], Fig. \ref{fig:cfx}c) and compromise trials (mean difference = 0.022, HDI$_{95}$  = [0.014, 0.030], mean $d$ = 0.90, HDI$_{95}$  = [0.53, 1.30], Fig. \ref{fig:cfx}f). Targets and competitors also received greater relative dwell time than decoys in both trial types.

To better understand participants' eye movements over the course of the trial, we performed additional exploratory analyses of gaze location and transitions. We found an increasing association between gaze and choice over the trial, and longer gaze towards targets, even accounting for choice. Information search occurred both within and between alternatives, with slightly more transitions within alternatives. We refer to \nameref{sup:eye-movement-analyses} for additional details on these analyses.

@@

% Fig. 3: Model comparison results, predicted gaze-choice association RST prediction of winning model
\begin{figure}[bt!]
\begin{centering}
\includegraphics[scale=1]{../figures/3-model-comparison.pdf}
\caption{\textbf{Model comparison and predictions.} \textbf{(a)} The gaze-dependent leaky accumulator (GLA) provided the best fit (lowest mean BIC) across participants, followed by the dynamic gaze baseline model, MDFT, EU, and the static gaze baseline model. The dashed line indicates the BIC of the random choice baseline model. Violin plots show kernel density estimates of distributions of individual values. Box plots mark lower and upper quartiles and median. Whiskers extend from first and last datum within 1.5 times the interquartile range from lower and upper quartiles, respectively. Values outside this range are indicated by open circles. \textbf{(b)} The GLA fitted most (36 of 40, 90\%) participants best, with a protected exceedance probability of 1 (inset). \textbf{(c-h)} Observed and model-predicted probability of choosing the target \textbf{(c, d)}, competitor \textbf{(e, f)}, or decoy \textbf{(g, h)} alternatives in attraction \textbf{(c, e, g)} and compromise trials \textbf{(d, f, h)} as a function of relative dwell time advantage. Relative dwell time advantage was computed as relative dwell time towards an alternative minus the mean relative dwell time to all other alternatives. White bars and error bars show mean $\pm$ s.e. observed data from even-numbered trials. Model predictions (coloured lines) are based on 50 simulations of each odd-numbered trial. \textbf{(i, j)} Observed and predicted RST of the best-fitting GLA for attraction \textbf{(i)} and compromise \textbf{(j)} trials. Each circle represents one participant. The winning model’s predicted context effect sizes correlated significantly with the observed ones. Strong context effects, however, were underestimated, as indicated by the reduced slopes.}
\label{fig:modelcomparison}
\end{centering}
\end{figure}

@@

where all $X_i(0)=0$. Similar to the aDDM, the $\theta$ parameter ($0 \le \theta \le 1$) controls discounting of unattended alternative utilities. The $\lambda$ parameter ($0 \le \lambda \le 1$) controls the strength of the accumulation leak. We computed choice probabilities from this model by applying the soft-max choice rule (Eq. (\ref{eq:softmax})) over the final accumulator values $X_i^{GLA}$.

@@

Across the group, all models performed better than the random baseline model (Fig. \ref{fig:modelcomparison}a). The GLA had the lowest BIC (mean $\pm$ s.d. = 230.63 $\pm$ 66.80), followed by the dynamic gaze baseline model GB$_{\text{dyn}}$ (mean $\pm$ s.d. = 302.46 $\pm$ 86.15), MDFT (mean $\pm$ s.d. = 345.77 $\pm$ 82.44), EU (mean $\pm$ s.d. = 360.57 $\pm$ 72.56) and the static gaze baseline GB$_{\text{stat}}$ (mean $\pm$ s.d. = 414.81 $\pm$ 36.81). Simulating choices from the models using maximum-likelihood estimates, the proportion of correctly predicted choices was 74.2\% for GLA, 62.8\% for GB$_{\text{dyn}}$, 57.9\% for MDFT, 53.8\% for EU and 45.7\% for GB$_{\text{stat}}$ (see Supplementary Fig. \ref{fig:model-pred-probs} for distributions of model-predicted choice probabilities). Note, that target and competitor options (and decoys in compromise trials) were designed to be closely matched in value, resulting in trials with high choice difficulty, limiting overall model performance. On the individual level, based on lowest BIC scores, the majority (36 of 40, 90\%) of participants were best described by GLA. Three participants (7.5\%) were best described by MDFT, and one by the dynamic gaze-baseline model (Fig. \ref{fig:modelcomparison}b). Protected exceedance probabilities \parencite{rigoux2014BayesianModelSelection}, which measure the likelihood that model is more frequent than all others, unambiguously identified GLA as the most likely model (pXP$_{\text{GLA}}$ = 1, Fig. \ref{fig:modelcomparison}b inset).

The estimates for GLA’s gaze-discount parameter $\theta$, which describes the degree to which alternative’s values are attenuated while not fixated, indicate that participants exhibited a moderate gaze-discount on average: $\theta$ estimates ranged from 0.13 (strong gaze-discount) to 0.95 (almost no gaze-discount), with mean $\pm$ s.d. = 0.69 $\pm$ 0.18. Estimates for the leak parameter ranged from 0.08 (almost no leak) to 0.65 (moderately strong leak), with mean $\pm$ s.d. = 0.29 $\pm$ 0.20. Individual parameter estimates of all GLA parameters are plotted in Supplemental Fig. \ref{fig:gla-estimates} and summarised in Supplemental Table \ref{tab:gla-estimates}.

Given that MDFT outperforms utility-based models when choices are influenced by context \parencite{berkowitsch2014RigorouslyTestingMultialternative}, we tested whether model fit of MDFT (relative to GLA) was higher for participants with higher RST. In other words: Does MDFT perform better for stronger context effects? Overall, 7 out of 9 participants with an RST above 0.7 in attraction trials were clearly best described by GLA. While we found that the relative superiority of GLA over MDFT decreased with strong attraction effects, this was not the case for compromise trials (Supplemental Fig. \ref{fig:dbic-rst}), suggesting that some features of MDFT might help capture strong attraction effects.

As an additional indicator of model fit, we tested whether the models could quantitatively reproduce the observed positive association of gaze and choice (Supplemental Fig.s \ref{fig:dwell-data-timebinned}d, j, \ref{fig:dwell-regression-weights}). Specifically, following previous work \parencite{krajbich2010VisualFixationsComputation,krajbich2011MultialternativeDriftdiffusionModel}, we inspected the model-predicted probability of choosing an alternative as a function of its relative dwell time advantage: the difference in relative dwell time towards an alternative minus the mean relative dwell time to other alternatives. The probability of choosing an alternative generally increased with its relative dwell time advantage (Fig. \ref{fig:modelcomparison}c-h), except for dominated decoys in attraction trials, which were not chosen even if looked at longer than other alternatives (Fig. \ref{fig:modelcomparison}g).

All gaze-dependent models were able to capture this positive association. Note, however, that they also predicted choices of dominated decoys in attraction trials, if decoys had a large dwell time advantage (Fig. 3g). In this case, GLA performed better than GB$_{\text{dyn}}$ and GB$_{\text{stat}}$, as it predicted fewer decoy choices. However, MDFT and EU generally could not capture the empirical association of gaze and choice; they predicted too many choices of alternatives with negative dwell time advantage, and too few choices of alternatives with positive dwell time advantage. 
Nor did they predict choices of dominated decoys, even if the decoy was looked at longer.

Finally, we investigated the ability of the best-performing model to predict individual differences in context effect strengths. Therefore, we predicted choices from the fitted GLA model and correlated the resulting RST with the observed data. Predicted RST correlated significantly with observed ones in attraction ($r$ = 0.76, HDI$_{95}$ = [0.62, 0.90], Fig. \ref{fig:modelcomparison}i) and compromise ($r$ = 0.82, HDI$_{95}$ = [0.72, 0.92], Fig. \ref{fig:modelcomparison}j) trials. However, the model underestimates large deviations from RST = 0.5, suggesting that its gaze-discount mechanism can capture the qualitative pattern of context effects but not their expression in participants with extreme RSTs.

Inspection of predicted choice probabilities (Supplemental Fig. \ref{fig:model-pred-probs}) shows that, on average, GLA predicted high probabilities for the empirically chosen alternative (indicating good overall fit), but comparable proportions of target and competitor choices (resulting in reduced RST). Other models, like MDFT, predicted larger differences between target and competitor choices for some participants, but assigned lower probability to empirically chosen alternatives (resulting in overall inferior fit).
